---
title: "Are Public Demonstrations a Driving Force for Politics?"
author: "Shengpei Wang"
date: "May 14, 2019"
output: github_document
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = F, message = FALSE, warning = FALSE)
```
```{r load_data}
load("figures_and_data_for_markdown.RData")
```

# Is protesting useful?

Public demonstrations have a long history of driving political changes around the globe, such as the Civil Rights Movement in the US, the Indian Independence Movement in India, and The Defiance Campaign in South Africa. The effects of demonstrations have been studied in detail in some specific cases, such as the Tea Party movement in the US (Madestam *et. al* 2011). However, whether demonstrations are generally effective as a political force is still unclear. 

A complicating factor affecting the effectiveness of demonstrations is the political environment in which demonstrations happen. We may expect that people are more inclined to attend public demonstrations in democratic systems, because attending demonstrations might be seen as an effective political action. Alternatively, demonstrations could be more common in authoritarian regimes, because alternative political actions may not be available or perceived to be effective, such as in the case of the Arab Spring Uprising in Egypt. Given the increasing social impacts of demonstrations through social media, it is more urgent than ever to understand the prevalence and effectiveness of public demonstrations in different political systems. 

I propose to investigate the prevalence and effectiveness of demonstrations by integrating information from two large public datasets: The Rulers, Elections, and Irregular Governance Dataset (REIGN; Bell 2016) and the Mass Mobilization in Autocracies Database (MMAD; Weidmann *et. al* forthcoming). The REIGN data records the leader and regime information from 201 countries between 1950-current, while the MMAD data records incidents of political demonstrations in 76 countries between 2002 and 2015. By combining both data sources, I plan to examine both the political environments and the political outcomes of public demonstrations across the world,

I have four major questions/objectives of interest: 

* **What types of public demonstrations happen in democratic regime versus autocratic regimes?**
    + Are pro-regime demonstrations more or less common than anti-regime demonstrations?
    + Does the frequency of pro- vs anti-regime demonstration differ depending on the types of regimes?
* **What types of public demonstrations have the largest impact on society?**
    + Which demonstrations do people attend?
    + Do elections change the size of demonstrations?
    + Are demonstration sizes or locations correlated with different election results?
    + Are demonstration sizes or locations correlated with changes in the leadership?
* **Which is the cause: Do governments respond to demonstrations or do people demonstrate in response to governments?**
    + What kinds of demonstrations happen before an election or leadership change?
    + What kinds of demonstrations happen after an election or leadership change?
    + Are elections or leadership changes more likely to happen before or after demonstrations?
    + Are authoritarian leaders more likely to step down after demonstrations?
* **Can we predict the political changes given the current political environment and the prevalence of demonstrations around the world?**
    + Build a machine learning algorithm that predicts whether elections or leadership changes will happen given the current government regimes and the type and amount of demonstrations.
    + Build a machine learning algorithm that predicts what kinds of demonstrations will happen given the political environment.

Answering these questions will help us understand roles of public demonstrations in driving political changes. The results can help inform citizens when making political decisions regarding activism. The insights can further help policy makers understand, predict, and possibly prepare for political trajectories of different countries around the world.

# Methods

I will investigate these questions by joining the two datasets and investigate the correlations among the different factors in each question. There are two major ways I plan to structure the data. One way is to examine the political environments and demonstrations that occur during the same time period, and this will give us whether these factors are correlated in different countries. An alternative way is to examine the temporal order of events that represents political changes and demonstrations, and I will data in this structure to untangle the causal factors as well as to build the predicative models. 

I will use a variety of statistical and machine learning approaches to test alternative hypotheses and use different visualizions to communicate the results. I will use xi-squared test and logistic regression for many of the questions about correlations. I will use linear mixed models and clustering analysis to build the predicative models.

I plan to extend these analyses to include other important factors such as the economic, social, and religious situations of each country over time. These factors can affect both the political environment and people's willingness and desire to demonstrate. I have obtained data from the Human Mortality Database that contains historical demographic information of different countries. I have also obtained and the World Religion Data that details the religious composition of each country (Maoz and Henderson 2013). Integration of these additional data sources will allow me to obtain a more holistic view of the drivers of political changes around the world. 


# Data Sources and description:
Both datasets are publicly available on their websites:

*  REIGN: https://oefresearch.org/datasets/reign
*  MMAD: https://mmadatabase.org/

The REIGN dataset. I downloaded the most recent full dataset from their archive.

This is a fairly complicated dataset, and I will be focusing on only some of the variables for the preliminary exploration. Each row records the leader and political environment of a single month in each country.

* "ccode" is the country code based on the The Correlates of War Project
* "country" is the name of the country
* "leader" is the name of the leader during the recorded time
* "elected" represents whether the leader is elected to office (1) or not (0).
* "tenure_months" is the length of the tenure of the leader for the month of record
* "government" is the type of political regime of the country at the month of record.

```{r data_description_REIGN}
data_REIGN_sample
```

Look at the MMAD dataset. I obtained the data describing the incidents of each recorded event.

* "cowcode" is the unique country code also based on the The Correlates of War Project. 
* "asciiname" is the city where the demonstration happened.
* "event_date" is the date when the demonstration happened.
* "side" represents whether the demonstration is pro-regime (0), anti-regime (1) or domestic public/non-government (3). Domestic demonstrations are rare and does not affect my questions, so I will removed them for my analyses.
* "mean_avg_numparticipants" represent the best estimate of the size of the demonstrations.

```{r data_description_NNAD}
data_MMAD_sample
```

# Popularity of mass demonstrations
## Demonstrations are in the news all the time, but how common are they across the world?

I examined the prevalence of demonstrations by tallying the number of months when demonstrations did or did not happen for each country during 2003-2015. I further sectioned the data by the type of regime, because it could be a major driving factor for whether people decide to participate in public demonstrations. 

```{r fig1}
plot_freq_dem
```

Demonstrations were much more frequent in all of the autocracies than the two democratic regimes. This shows that people chose to use demonstration as a political tool more often in autocracies. 

But were people actually happier with their government in democratic systems than autocratic systems? This pattern could be driven by more pro-regime demonstrations in autocratic systems. I further explored the prevalence of demonstrations that were pro-regime vs anti-regime across different regimes.

```{r fig2}
plot_prop_dem
```

Most demonstrations turn out to be anti-regime regardless of the types of regimes. Additionally, pro-regime demonstrations tended to happen concurrently with anti-regime demonstrations. This makes sense that there is no political need to express one's support for a regime unless it is challenged. 

## Do patterns of demonstration depend on the legitimacy of a leader?

The previous section shows that people were happier with democratic regimes. Was this driven by the fact that the leaders were elected? I investigated this question by looking at both numbers and sizes of demonstrations with legitimate vs illegitimate leaders. I quantified the legitimacy of leaders as whether they were elected to office.

```{r fig3}
plot_anti_election
```

There were more anti-regime demonstrations when a leader was not elected to office. Demonstrations against illegitimate leaders were also better attended. Even though the majority of demonstrations are anti-regime, the legitimacy of the leaders did matter. Does the same pattern hold for pro-regime demonstrations?

```{r fig4, fig.width = 8}
plot_pro_election
```

Concordant with the previous findings, illegitimate leaders were less supported by pro-regime demonstrations, which were also less popular compared to with legitimate leaders. These results demonstrate that greater prevalence of demonstrations in autocratic systems was indeed partially correlated with the legitimacy of the leaders. 

One interesting outlier was the party-personal regime, where there were more pro-regime demonstrations than all other regimes and that illegitimate leaders enjoyed greater support through demonstrations. I explored this pattern further by look into each country separately.

```{r fig5}
plot_cuba_nkorea
```

## Do demonstrations drive regime change or happen in response to regime change?

In the previous sections, I found that demonstration was a popular political tool to protest against illegitimate leaders or autocratic regimes. But were demonstrations effective at driving political changes? One way to answer this question is to look at the timing of events, did demonstrations happen before or after a leader took office. 
```{r fig6, fig.width = 8}
plot_dem_dots
```

Each point in the figure above represents one public demonstration. Even though the prevalence of demonstration differed amongst regime types, more demonstrations tended to happen later in a leader's tenure, thus before rather than after leadership changes.   

## Bibliography
Madestam, A., Shoag, D., Veuger, S. and Yanagizawa-Drott, D., 2013. Do political protests matter? evidence from the tea party movement. The Quarterly Journal of Economics, 128(4), pp.1633-1685.

Bell, Curtis. 2016. The Rulers, Elections, and Irregular Governance Dataset (REIGN). Broomfield, CO: OEF Research. Available at oefresearch.org

Weidmann, Nils B. and Espen Geelmuyden Rød. The Internet and Political Protest in Autocracies. Chapter 4. Oxford University Press, forthcoming.
